Modal Abstractions for Virtualizing Memory Addresses

### 2023-04-11 programlogic.tex
+

  Definition vProp  $\Sigma$ : bi := word 64 $\rightarrow$_{b} iPropI  $\Sigma$.

@@

\subsection{Points-To Assertions}
\label{sec:pointsto}
\todo[inline]{Colin says: Current limit of \emph{thorough} revisions}
\todo[inline,color=red]{Colin says: This section at a glance looks on the right track, but none of the assertions are actually
written as vProps, they're all still iProps. I can fix that as I go}
\todo[inline]{Colin says: There's really no need to discuss the 52-12 split in the physical pointsto definition.
That level of detail of the operational semantics \emph{is} too detailed for the paper (but should be mentioned in the
artifact docs). I also don't know what the `ent' and `rtv' and similar are in those split physical pointstos}
We abstract physical memory addressing and registers naming with well-known separation logic assertion \textit{points-to} assertions.
\begin{enumerate}
\item Physical address points-to, $\pfpointsto\locsf\vpts\qfrac\ppts$
\item Register points-to, $\pfpointsto\rg\rv\qfrac\rpts$
\end{enumerate}
\paragraph{Register points-to} The assertion $\pfpointsto\rg\rv\rpts\qfrac$ ensures the ownership of the register $\rg$ naming the register value $\rv$. The fraction $\qfrac$ with value 1 asserts the unique ownership of the register mapping, and grants update permission on it, otherwise, any value $0 < \qfrac <1$ represents partial ownership granting readonly permission on the mapping.

@@

Given the definition of physical page-pointsto assertion and the root address of virtual-address space kept in the control register \lstinline{cr3} as shown in Figure \ref{fig:pagetable},  one can build the physical address-translation for a virtual address (e.g. \textsf{va}) via abstracting the L4-L1 table traversal as the following:
\begin{itemize}
  \item Page Map Level-4 Translation (L4): Performs the first level address translation to get the PML4 entry (PML4E) in Figure \ref{fig:pagetables} by using the masks (\textsf{va}$|^{52}$(\lstinline{cr3}) to get the starting address of the PML4 table (PLM4T) and using the offset mask \textsf{va}$|^{12}$(\lstinline{cr3}). It is worth noting that our offset indices are 12 bits as all physical memory pointsto assertions has to satisfy being \textsf{\textit{aligned}}, e.g. [47-39] plus 3 alignment bits in Figure \ref{fig:pagetables}
    \[ \begin{array}{l}
      \hbox{(\TirNameStyle{{L4translate}})} \quad
      \exists \entryf \ldotp \nfpointsto{\mask\vaddr\ft\rtv}{\mask\vaddr\tw\rtv}\entryf\qone\naddr
      \end{array}
      \]
  \item Page Directory Pointer Level Translation (L3): Performs the second level address translation to get the PDP entry (PDPE) in Figure \ref{fig:pagetables} by using the masks(\textsf{va}$|^{52}$(\textsf{PLM4E}) to get the starting address of the PDP table (PDPT) and using the offset mask \textsf{va}$|^{12}$(\textsf{PLM4E})
    \[\begin{array}{l}
    \hbox{(\TirNameStyle{{L3translate}})} \quad
    \exists \entrytr \ldotp \nfpointsto{\mask\vaddr\ft\entryf}{\mask\vaddr\tw\entryf}\entrytr\qtwo\naddr
    \end{array}
    \]
  \item Page Directory Table Level Translation (L2): Performs the second level address translation to get the PD entry (PDE) in Figure \ref{fig:pagetables} by using the masks(\textsf{va}$|^{52}$(\textsf{PDPE}) to get the starting address of the PD table and using the offset mask \textsf{va}$|^{12}$(\textsf{PDPE})
    \[ \begin{array}{l}
            \hbox{(\TirNameStyle{{L2translate}})} \quad 
       \exists \entrytw \ldotp \nfpointsto{\mask\vaddr\ft\entrytr}{\mask\vaddr\tw\entrytr}\entrytw\qthree\naddr
      \end{array}
      \]
  \item Page Table Level Translation (L1): Performs the second level address translation to get the PT entry (PTE) in Figure \ref{fig:pagetables} by using the masks(\textsf{va}$|^{52}$(\textsf{PDE}) to get the starting address of the Page table (PT) and using the offset mask \textsf{va}$|^{12}$(\textsf{PDE})
    \[ \begin{array}{l}
      \hbox{(\TirNameStyle{{L1translate}})} \quad
      \exists \entryo \ldotp \nfpointsto{\mask\vaddr\ft\entrytw}{\mask\vaddr\tw\entrytw}\entryo\qfour\naddr
      \end{array}
      \]
  \item Page Address Level Translation: Performs the second level address translation to get the physical page address in Figure \ref{fig:pagetables} by using the masks(\textsf{va}$|^{52}$(\textsf{PTE}) to get the starting address of the physical page address table and using the offset mask \textsf{va}$|^{12}$(\textsf{PTE})
    \[ \begin{array}{l}
      \hbox{(\TirNameStyle{{PageLebelAccess}})} \quad
      \exists \vpage \ldotp \nfpointsto{\mask\vaddr\ft\entryo}{\mask\vaddr\tw\entryo}\vpage\qfracone\naddr
      \end{array}
      \]

\end{itemize}
\subsection{An Overly-Restrictive Definition for Virtual Memory Addressing}
One could give a naive logical definition to the virtual mememory addressing as conjuctions of each level translation 
(Figure \fref{fig:strongvirtualpointsto}). The very first assurance our logical constructions need to be able assert is 
the \textit{existence} of the page table walk. Our naive attempt shown in Figure \ref{figure:strongvirtualpointsto}, 
we see that a virtual address with a certain fractional permissions at each level of the page-table walk gives us sound 
assurance of the path existence reaching to a page table entry, consequently, with a full-ownership of page address, 
we can obtain and update the value mapped value at this physical page address through our virtual memory address.

\todo[inline]{Cite Kolanski here; not all of them, but some actually defined it exactly this way, for the current
address space.}

\begin{figure*}
  \begin{lstlisting}[language=Coq]
 Definition $\pfpointsto\locsf\vpts\qfrac\ppts$ :vProp $\Sigma$ := pmapsto_def_v (pfn w) (page_off w) q v.
 Definition $\vaddr\mapsto_{\textsf{t}}\{\textsf{q}\}\vpage$ :vProp $\Sigma$ := 
  $\exists_{\entryf \;, \entrytr \;, \entrytw \;,\entryo} \ldotp$  $\ulcorner$ aligned $\vaddr \urcorner \ast$ L4_L1_PointsTo($\rtv,\entryf,\entrytr,\entrytw,\entryo$) $\ast$ addr_L1$(\vaddr,\entryo) \mapsto_{p} \vpage$.
 Definition L4_L1_PointsTo (rtv maddr entry entry0 entry1 pte :word 64) : vProp $\Sigma$ :=
  ((l4M52 maddr rtv) \$\sim$\ (l4off maddr rtv) $\mapsto_{a}$ {q1} entry  $\ast$
  (l3M52 maddr entry) \$\sim$\ (l3off maddr entry)  $\mapsto_{a}$ {q2} entry0 $\ast$ 
  (l2M52 maddr entry0) \$\sim$\ (l2off maddr entry0) $\mapsto_{a}$ {q3} entry1 $\ast$
  (l1M52 maddr entry1) \$\sim$\ (l1off maddr entry1) $\mapsto_{a}$ {q4} pte ).
\end{lstlisting}
\caption{A Strong Virtual Points-to Relation}
\todo[inline,color=red]{Colin says: shouldn't that fraction be a q, not a 1, in the second line?}
  \label{fig:strongvirtualpointsto}
\end{figure*}

\subsection{Aliasing/Sharing Physical Pages}
  \label{sec:sharingpages}  
  One might have already observed that the virtual points-to definition shown in Figure \ref{fig:strongvirtualpointsto} 
  is too strong to specify some operations updating a page table entry.
   To speak more concretely, the fractional permissions on the page-table-walk assertion ($\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ in Figure \ref{fig:strongvirtualpointsto}) 
  ensure fractional ownership of the four physical page table entries accessed by the hardware during address translation,
  as well as the physical memory holding the data of the virtual points-to.
  However, this formulation stores knowledge of the page table walk's details with ownership of the backing physical
  memory.
  Updating any of these mappings (e.g., move the page tables in physical memory, as in coallescing for
  superpages or hugepages)
  would require explicitly collecting all virtual points-to facts that traverse affected entries.
  It is preferable to permit the page tables themselves to be updated independently of the virtual points-to assertions,
  so long as those updates preserve the same virtual-to-physical translations.

>>>>

  Definition vProp  $\Sigma$ : bi := word 64 $\rightarrow_{b}$ iPropI  $\Sigma$.

@@

\subsection{Points-To Assertions}
\label{sec:pointsto}
\todo[inline]{Colin says: Current limit of \emph{thorough} revisions}
\todo[inline,color=red]{Colin says: This section at a glance looks on the right track, but none of the assertions are actually
written as vProps, they're all still iProps. I can fix that as I go}
\todo[inline]{Colin says: There's really no need to discuss the 52-12 split in the physical pointsto definition.
That level of detail of the operational semantics \emph{is} too detailed for the paper (but should be mentioned in the
artifact docs). I also don't know what the `ent' and `rtv' and similar are in those split physical pointstos}
We abstract physical memory addressing and registers naming with well-known separation logic assertion \textit{points-to} assertions.
\item Physical address points-to, $\pfpointsto\locsf\vpts\qfrac\ppts$
\item Register points-to, $\pfpointsto\rg\rv\qfrac\rpts$
\paragraph{Register points-to} The assertion $\pfpointsto\rg\rv\rpts\qfrac$ ensures the ownership of the register $\rg$ naming the register value $\rv$. The fraction $\qfrac$ with value 1 asserts the unique ownership of the register mapping, and grants update permission on it, otherwise, any value $0 < \qfrac <1$ represents partial ownership granting readonly permission on the mapping.

@@

\begin{figure*}
 Definition $\pfpointsto\locsf\vpts\qfrac\ppts$ :vProp $\Sigma$ := pmapsto_def_v (pfn w) (page_off w) q v.
 Definition $\vaddr\mapsto_{\textsf{t}}\{\textsf{q}\}\vpage$ :vProp $\Sigma$ := 
  $\exists_{\entryf \;, \entrytr \;, \entrytw \;,\entryo} \ldotp$  $\ulcorner$ aligned $\vaddr \urcorner \ast$ L4_L1_PointsTo($\rtv,\entryf,\entrytr,\entrytw,\entryo$) $\ast$ addr_L1$(\vaddr,\entryo) \mapsto_{p} \vpage$.
 Definition L4_L1_PointsTo (rtv maddr l4e l3e l2e l1e :word 64) : vProp $\Sigma$ :=
  (l4M52 maddr rtv) \$\sim$\ (l4off maddr rtv) $\mapsto_{a}$ {q1} l4e  $\ast$
  (l3M52 maddr entry) \$\sim$\ (l3off maddr entry)  $\mapsto_{a}$ {q2} l3e $\ast$ 
  (l2M52 maddr entry0) \$\sim$\ (l2off maddr entry0) $\mapsto_{a}$ {q3} l2e $\ast$
  (l1M52 maddr entry1) \$\sim$\ (l1off maddr entry1) $\mapsto_{a}$ {q4} l1e.
\caption{A Strong Virtual Points-to Relation}
\todo[inline,color=red]{Colin says: shouldn't that fraction be a q, not a 1, in the second line?}
  \label{fig:strongvirtualpointsto}
\end{figure*}
Given the definition of physical page-pointsto assertion and the root address of virtual-address space kept in the control register \lstinline{cr3} as shown in Figure \ref{fig:pagetable},  one can build the physical address-translation for a virtual address (e.g. \textsf{va}) via abstracting the L4-L1 table traversal as the following:
\begin{itemize}
  \item Page Map Level-4 Translation (L4): Performs the first level address translation to get the PML4 entry (PML4E) in Figure \ref{fig:pagetables} by using the masks (\textsf{va}$|^{52}$(\lstinline{cr3}) to get the starting address of the PML4 table (PLM4T) and using the offset mask \textsf{va}$|^{12}$(\lstinline{cr3}). It is worth noting that our offset indices are 12 bits as all physical memory pointsto assertions has to satisfy being \textsf{\textit{aligned}}, e.g. [47-39] plus 3 alignment bits in Figure \ref{fig:pagetables}
      $\hbox{(\TirNameStyle{{L4translate}})} \quad$ (l4M52 maddr rtv) \$\sim$\ (l4off maddr rtv) $\mapsto_{a}$ {q1} l4e 
 \item Page Directory Pointer Level Translation (L3): Performs the second level address translation to get the PDP entry (PDPE) in Figure \ref{fig:pagetables} by using the masks(\textsf{va}$|^{52}$(\textsf{PLM4E}) to get the starting address of the PDP table (PDPT) and using the offset mask \textsf{va}$|^{12}$(\textsf{PLM4E})
    $\hbox{(\TirNameStyle{{L3translate}})}$ (l3M52 maddr entry) \$\sim$\ (l3off maddr entry)  $\mapsto_{a}$ {q2} l3e
  \item Page Directory Table Level Translation (L2): Performs the second level address translation to get the PD entry (PDE) in Figure \ref{fig:pagetables} by using the masks(\textsf{va}$|^{52}$(\textsf{PDPE}) to get the starting address of the PD table and using the offset mask \textsf{va}$|^{12}$(\textsf{PDPE})
    $\hbox{(\TirNameStyle{{L2translate}})}$ (l2M52 maddr entry) \$\sim$\ (l2off maddr entry)  $\mapsto_{a}$ {q2} l2e
  \item Page Table Level Translation (L1): Performs the second level address translation to get the PT entry (PTE) in Figure \ref{fig:pagetables} by using the masks(\textsf{va}$|^{52}$(\textsf{PDE}) to get the starting address of the Page table (PT) and using the offset mask \textsf{va}$|^{12}$(\textsf{PDE})
    $\hbox{(\TirNameStyle{{L1translate}})}$ (l1M52 maddr entry) \$\sim$\ (l1off maddr entry)  $\mapsto_{a}$ {q2} l1e
  \item Page Address Level Translation: Performs the second level address translation to get the physical page address in Figure \ref{fig:pagetables} by using the masks(\textsf{va}$|^{52}$(\textsf{PTE}) to get the starting address of the physical page address table and using the offset mask \textsf{va}$|^{12}$(\textsf{PTE})
    $\hbox{(\TirNameStyle{{PageLabelAccess}})}$ addr_L1$(\vaddr$,\textsf{l1e}$) \mapsto_{p} \vpage$.
\end{itemize}
\subsection{An Overly-Restrictive Definition for Virtual Memory Addressing}
One could give a naive logical definition to the virtual mememory addressing as conjuctions of each level translation 
(Figure \fref{fig:strongvirtualpointsto}). The very first assurance our logical constructions need to be able assert is 
the \textit{existence} of the page table walk. Our naive attempt shown in Figure \ref{figure:strongvirtualpointsto}, 
we see that a virtual address with a certain fractional permissions at each level of the page-table walk gives us sound 
assurance of the path existence reaching to a page table entry, consequently, with a full-ownership of page address, 
we can obtain and update the value mapped value at this physical page address through our virtual memory address.

\todo[inline]{Cite Kolanski here; not all of them, but some actually defined it exactly this way, for the current
address space.}
\subsection{Aliasing/Sharing Physical Pages}
  \label{sec:sharingpages}  
  One might have already observed that the virtual points-to definition shown in Figure \ref{fig:strongvirtualpointsto} 
  is too strong to specify some operations updating a page table entry.
   To speak more concretely, the fractional permissions on the page-table-walk assertion ($\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ in Figure \ref{fig:strongvirtualpointsto}) 
  ensure fractional ownership of the four physical page table entries accessed by the hardware during address translation,
  as well as the physical memory holding the data of the virtual points-to.
  However, this formulation stores knowledge of the page table walk's details with ownership of the backing physical
  memory.
  Updating any of these mappings (e.g., move the page tables in physical memory, as in coallescing for
  superpages or hugepages)
  would require explicitly collecting all virtual points-to facts that traverse affected entries.
  It is preferable to permit the page tables themselves to be updated independently of the virtual points-to assertions,
  so long as those updates preserve the same virtual-to-physical translations.
